---
title: "Introduction to **R**"
subtitle: "Session 5"
author: "Statistical Consulting Centre"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output: pdf_document
---


```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = TRUE)

library(dplyr)
library(tidyr)

load("../../RData/Ex5.RData")
```


# 1. Box plot

Using `joined.long.df`, draw a box plot showing the distribution of `Mercury` for each of the levels of `Time`. The box plot should look exactly like 

```{r, fig.align='center'}
with(joined.long.df, plot(Calcium, Mercury,
                      xlab = "Day", cex.lab = 1.5,
                      ylab = "Mercury concentration",
                      main = "Mercury contaminations in Florida lakes"))

```

The box plot should have:

- labels on the x-axis,
- an informative label on the y-axis (not `Mercury`),
- a title, and
- labels that are large enough to read.


\newpage

# 2. Scatter plot
Using `joined.long.df`, draw a scatter plot which shows the relationship between `pH` and `Mercury` for each of the `Calcium` levels. The scatter plot should look the same as 

```{r, fig.align='center'}
with(joined.long.df, plot(pH, Mercury, 
     col = c(1,2,4)[Calcium], 
     pch = c(20, 22,3)[Calcium], 
     xlab = "pH value", ylab = "Mercury concentration"))

legend("topright", pch = c(20, 22, 3), 
       col = c(1, 2, 4),
       cex = 1,
legend = c("Low", "Medium", "High"))
```

Hint: 

- `Low Calcium` level: using `pch = 20` and `col = 1`.
- `Medium Calcium` level: using `pch = 22` and `col = 2`.
- `High Calcium` level: using `pch = 3` and `col = 4`.

\newpage

# 3. Bar plot

Using `joined.long.df`, draw a bar plot that shows the average `Mercury` concentration for the six combinations of `Calcium` and `pHtype`. The graph should look exactly the same as

```{r, fig.align='center'}
trt.mean <- with(joined.long.df, tapply(Mercury,
                            list(pHtype, Calcium), mean))

barplot(trt.mean, beside = TRUE, space = c(0.1, 1),
    legend.text= c("Acidity", "Neutral", "Alkalinity"),
            ylab = "Mean mercury concentration", 
            xlab = "Calcium level",
    args.legend = list(bty = "n", cex = 1.5))

abline(h = 0)
```

Draw the bar plot in the following steps:

- Calculate the mean `Mercury` level for each combination of `pHtype` and `Calcium`, using the `tapply()` function.
- Save the calculated means in a new object called `trt.mean`.
- Produce the barplot using `beside = TRUE`, putting a  small space  between the bars. (Reminder: the argument that does this is `space`).
- Add a legend.
- Line along horizontal axis.




