---
title: "Introduction to **R**"
subtitle: "Session 5 -- Basic Graphics"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output:
  beamer_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
    highlight: tango
    includes:
      in_header: ../../header.tex
---


```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(tidy = TRUE, message = FALSE, small.mar=TRUE)

options(width = 50)
library(dplyr)
library(tidyr)


```


# Two datasets from yesterday

```{r, eval = FALSE}
load("Rdata/Session5.Rdata")

ls()
```

```{r, echo = FALSE}
load("../../Rdata/Session5.Rdata")
ls()
```

- `load()`: reload **R** objects from a RData file. 
- `ls()`: list the names of all objects currently in **R**.
- `combined.df` contains the information at the patient level and
- `combined.long.df` contains the information at the measurment level. 
- You will be useing these two dataset for today.

# Designing graphs

There are 2 important considerations when designing a graph:

1. It should have something to say.

    + Decide what information you want your graph to display.

2. It should be easy to interpret.

    + Simplicity is key!
    + Sufficiently large fonts, all axes labelled, clear legends, etc.


# Some functions for drawing graphs

- Usually, we use `plot()` to create a graph and then `lines()`, `points()`, `text()`, and other commands to annotate it.
- `plot()` does appropriate things for different types of variables (see later examples).
- Variables of type `character` are not suitable for plotting (and analysis). Convert such variables to type `factor`.
- There are no "erase" or "undo" functions, so store your commands in an **R** script.

# `plot()`

Generic function for plotting of R objects.

```{r, fig.align='center'}
plot(combined.df)
```

# Box plots
```{r, error = TRUE}
with(combined.long.df, plot(Time, Cholesterol))
```


# Box plots

`Time` variable in `combined.long.df` has to be `factor`. 

```{r}

class(combined.long.df$Time)

combined.long.df$Time <- factor(combined.long.df$Time)

class(combined.long.df$Time)

```

# Box plots

```{r, fig.align='center'}
with(combined.long.df, plot(Time, Cholesterol))
```


# Box plots

```{r, eval = FALSE}
combined.long.df$Time <- factor(combined.long.df$Time,
                                levels = c("Baseline", "PreTrt", "PostTrt"))

with(combined.long.df, plot(Time, Cholesterol))
```

\vspace{-40pt}

```{r, fig.align='center', echo = FALSE}
combined.long.df$Time <- factor(combined.long.df$Time,
                                levels = c("Baseline", "PreTrt", "PostTrt"))


with(combined.long.df, plot(Time, Cholesterol))
```

#Formula interface
The plot() command can also be used this way:

```{r, eval = FALSE}
with(combined.long.df, plot(Cholesterol ~ Time))
```
\vspace{-40pt}

```{r, fig.align='center', echo = FALSE}
with(combined.long.df, plot(Cholesterol ~ Time))
```

# Boxplots

```{r, fig.align='center'}
with(combined.long.df, plot(Sex, Cholesterol))
```

# Axes need labels

Label axes with units.

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)"))
```


# `xlab` and `ylab`

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)"))
```

# Axes labels are too small

Label axes with units, large enough to read.

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5))
```

# `cex.lab = 1.5`

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5))
```

# Title

Sometimes you need a title.

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5, 
          main = "Cholesterol level for male and female"))
```


# `main`
```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5, 
          main = "Cholesterol level for male and female"))
```


# `boxplot()` works as well

```{r, eval = FALSE}
boxplot(Cholesterol~Sex, data = combined.long.df,
          xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5, 
          main = "Cholesterol level for male and female")
```
\vspace{-25pt}
```{r, fig.align='center', echo = FALSE}
boxplot(Cholesterol~Sex, data = combined.long.df,
          xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5, 
          main = "Cholesterol level for male and female")
```

# Smoking group by Age group

Examine the data via two-way frequency table. 

```{r} 
with(combined.df, table(Smoke.group, Age.group))

```

Note that we are using data at the **patient level**, i.e. `combind.df`. 

# Smoking group by Age group

Then, convert the counts to percentages.

```{r, tidy = FALSE}
smoke.age <- with(combined.df, 100 * 
                    prop.table(table(Smoke.group, Age.group), 
                               margin =  2))

round(smoke.age, 2)

```

# `barplot`
    
```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, xlab = "Age group by years", 
        ylab = "Percentage")

```
    
# `barplot`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, xlab = "Age group by years",
        ylab = "Percentage")

```   
    
# Side-by-side `barplot`
    
```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage")

```
    
# `beside = TRUE`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, xlab = "Age group by years", ylab = "Percentage")

```      
    
# Line along horizontal axis?
    
```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage")
abline(h = 0)
```
    
# `abline(h = 0)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", ylab = "Percentage")
abline(h = 0)
```  

# Vertical axis higher than bars?

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100))
abline(h = 0)
```

# `ylim = c(0, 100)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, xlab = "Age group by years", ylab = "Percentage", ylim = c(0, 100))
abline(h = 0)
```  

# Adding space between bars 

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5))
abline(h = 0)
```

# `space = c(0.2, 1.5)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5))
abline(h = 0)
```  

# Shading with lines

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 10))
abline(h = 0)
```

# `density = c(0, 10)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 10))
abline(h = 0)
```  

# Shading with lines
The default colour for shading lines is pale while! Change to black using the `col` argument. 

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 10), 
        col = "black")
abline(h = 0)
```  

# `col = "black"`
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 10), 
        col = "black")
abline(h = 0)
```  

# Color-filled bars? 

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        col = c("blue", "green"))
abline(h = 0)
```

# `col = c("blue", "green")`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        col = c("blue", "green"))
abline(h = 0)
```  

# Legend? 

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE)
abline(h = 0)
```

# `legend.text = TRUE`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE)
abline(h = 0)
``` 

# Further improvements

- Larger title for axes
- Get rid of the box drawn outside the legend
- Larger legend

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE, 
        args.legend = list(bty = "n", cex = 1.3))
abline(h = 0)
```

# ` args.legend = list(bty = "n", cex = 1.3)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE, 
        args.legend = list(bty = "n", cex = 1.3))
abline(h = 0)
``` 

# Histgrams 
```{r, tidy = FALSE, fig.height = 5, fig.width = 7, fig.align='center'}
hist(combined.long.df$Cholesterol)
```

# Fewer bins?  
```{r, tidy = FALSE, fig.height = 5, fig.width = 7, fig.align='center'}
hist(combined.long.df$Cholesterol, breaks = 5)
```

# More bins?  
```{r, tidy = FALSE, fig.height = 5, fig.width = 7, fig.align='center'}
hist(combined.long.df$Cholesterol, breaks = 40)
```

# Use meaningful axis labels
```{r, tidy = FALSE, eval = FALSE}
hist(combined.long.df$Cholesterol, main = NULL, 
     xlab = "Cholesterol level (mg/100ml)")
```
\vspace{-30pt}
```{r, fig.height = 5, fig.width = 6, fig.align='center', echo = FALSE}
hist(combined.long.df$Cholesterol, main = NULL, 
     xlab = "Cholesterol level (mg/100ml)")
```

# Scatter plots

- Used to display paired quantitative (numeric) data, e.g. age and Cholesterol level. 
- Make sure they are numeric. 
- Use `plot()` function. 

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, plot(Age, Cholesterol))

```

# Scatter plots

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, plot(Age, Cholesterol))

```

# `type =` 

type = controls how data are plotted.

- `type = "p"`: points, this is the default,
- `type = "l"`: lines,
- `type = "b"`: both lines and points,
- `type = "n"`: no plotting, which can set up a plot for latter additions.

# Plotting character

We will usually use different plotting character for different groups.
Argument `pch =` in `plot()` or `points()` is used to change the plotting character:

- `pch = 19`: solid circle,
- `pch = 20`: smaller solid circle,
- `pch = 21`: filled circle,
- `pch = 22`: filled square,
- `pch = 23`: filled diamond,
- `pch = 24`: filled triangle point up,
- `pch = 25`: filled triangle point down.

# Scatter plots

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, plot(Age, Cholesterol,
              col = c("black", "red")[Sex],
              pch = c(19, 24)[Smoke.group]))

legend("topright", pch = c(19, 19, 24, 24),
       col = c("black", "red", "black", "red"),
       legend = c("Non-Smokers:Female", "Non-Smokers:Male",
                  "Smokers:Female", "Smokers:Male"))
```

# Scatter plots

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, plot(Age, Cholesterol,
              col = c("black", "red")[Sex],
              pch = c(19, 24)[Smoke.group]))

legend("topright", pch = c(19, 19, 24, 24),
       col = c("black", "red", "black", "red"),
       legend = c("Non-Smokers:Female", "Non-Smokers:Male",
                  "Smokers:Female", "Smokers:Male"))
```

# Scatter plots

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, plot(Age, Cholesterol,
              col = c("black", "red")[Sex],
              pch = c(19, 24)[Smoke.group]))

legend("topleft", pch = c(19, 19, 24, 24),
       col = c("black", "red", "black", "red"),
       legend = c("Non-Smokers:Female", "Non-Smokers:Male",
                  "Smokers:Female", "Smokers:Male"))
```

# Scatter plots

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, plot(Age, Cholesterol,
              col = c("black", "red")[Sex],
              pch = c(19, 24)[Smoke.group]))

legend("topleft", pch = c(19, 19, 24, 24),
       col = c("black", "red", "black", "red"),
       legend = c("Non-Smokers:Female", "Non-Smokers:Male",
                  "Smokers:Female", "Smokers:Male"))
```


# Graphical Parameters

```{r, eval = FALSE, tidy = FALSE}
?par
op <- par(mfrow = c(2, 2))       
  with(combined.long.df, boxplot(Cholesterol~Age.group,
      xlab = "Age Group", ylab = "Total score"))
  barplot(smoke.age, beside = TRUE, ylab = "Percentage",
      xlab = "Age group in years", ylim = c(0, 100),
      space = c(0.2, 1.5), legend.text = TRUE)
      abline(h = 0)
  hist(combined.long.df$Cholesterol)
  with(combined.long.df, plot(Age, Cholesterol))
## At end of plotting, reset to previous settings:
par(op)

```

# Graphical Parameters

```{r, echo=FALSE, fig.align='center'}
op <- par(mfrow = c(2, 2))       
  with(combined.long.df, boxplot(Cholesterol~Age.group,
      xlab = "Age Group", ylab = "Total score"))
  barplot(smoke.age, beside = TRUE, ylab = "Percentage",
      xlab = "Age group in years", ylim = c(0, 100),
      space = c(0.2, 1.5), legend.text = TRUE)
      abline(h = 0)
  hist(combined.long.df$Age)
  with(combined.long.df, plot(Age, Cholesterol))
## At end of plotting, reset to previous settings:
par(op)

```

# Summary

- Box plots
- Bar plots
- Histgram
- Scatter plots















































    
