---
title: "Introduction to **R**"
subtitle: "Session 6 -- Advanced Graphics"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output:
  beamer_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
    highlight: tango
    includes:
      in_header: ../../header1.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, message = FALSE, small.mar=TRUE)

options(width = 70)
library(dplyr)
library(tidyr)
library(knitr)

load("../../Rdata/Session5.Rdata")
```

# `ggplot2` package
- Documentation: 
[http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)

- recommended reading "The Layered Grammar of Graphics":
[http://vita.had.co.nz/papers/layered-grammar.pdf](http://vita.had.co.nz/papers/layered-grammar.pdf)

- Load `ggplot2` package
```{r}
library(ggplot2)
```

# Creating a new ggplot

To define the coordinate system, we need to provide 3 pieces of information:

1. Dataset (`combined.long.df`)
2. Which variable to plot on the x axis (`Age`)
3. Which variable to plot on the y axis (`Cholesterol`)

```{r, eval = FALSE}
ggplot(data = combined.long.df,
       mapping = aes(x = Age, y = Cholesterol))
```

# Create a new ggplot

- Initialising a ggplot object.
```{r, fig.height=2, fig.width=3, warning=FALSE, fig.align='center'}
ggplot(data = combined.long.df,
       mapping = aes(x = Age, y = Cholesterol))
```

# Create a new ggplot

- Initialising a ggplot object.
```{r, eval=FALSE}
ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y = Cholesterol))
```
There are three common ways to invoke ggplot:

- `ggplot(combined.long.df, aes(x, y, <other aesthetics>))`
- `ggplot(combined.long.df)`
- `ggplot()`

# Create a new ggplot
- assign this ggplot object to a variable
```{r, fig.height=2, fig.width=3, warning=FALSE, fig.align='center'}
g <- ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y =  Cholesterol))
```

# Create a Scatter plot
We will tell **R** that we want to add (using `+`) points (`geom_point()`) at the coordinates to create a scatter plot.
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
g + geom_point()
```
\vspace{-20pt}
- `geom`, short for geometric object, describes the type of plot that you will produce.

# Create a Scatter plot
- Note that here are three common ways to invoke ggplot: 
```{r, eval=FALSE, tidy = FALSE}
ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y =  Cholesterol)) +
  geom_point()
ggplot(data = combined.long.df) + 
  geom_point( mapping = aes(x = Age, y =  Cholesterol))
ggplot() + 
  geom_point(data = combined.long.df, 
             mapping = aes(x = Age, y =  Cholesterol))
```
- Always check the documentation, `?geom_point`, for which aesthetics can be used. 

# First method is recommended
```{r, fig.height=2, fig.width=3, warning=FALSE, fig.align='center'}
g <- ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y =  Cholesterol))
g + geom_point()
```

# Control `colour` and `shape`
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
g + geom_point(aes(colour = Sex, shape = Smoke.group))
```

# Control `colour` and `shape`
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center', echo = FALSE}
g + geom_point(aes(colour = Sex, shape = Smoke.group))
```
- Always check the documentation, `?geom_point`, for which aesthetics can be used.
- Note the missing values in the legend labeling.


# Modify axis, legend, and plot labels
```{r, fig.height=2.25, fig.width=4, warning=FALSE, fig.align='center'}
g <- g + geom_point(aes(colour = Sex, shape = Smoke.group))+
   labs(title = "Cholesterol level versus Age",
   x = "Age", y = "Cholesterol level (mg/100ml)")
g
```

# `theme()` controls non-data components of the plot
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center', tidy=FALSE}
g + theme(plot.title = element_text(size=8, hjust = 0.5),
        axis.title = element_text(size=8),
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8) )
```

# Create a Scatter plot
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center', tidy=FALSE}
g <- ggplot(
  data = na.omit(combined.long.df[,
      c("Age", "Cholesterol", "Sex", "Smoke.group")]), 
  mapping = aes(x = Age, y =  Cholesterol, 
                color = Sex, shape = Smoke.group)) + 
  geom_point() +
  labs(title ="Cholesterol level versus Age",  
        x = "Age", y = "Cholesterol level (mg/100ml)")
```

- `ggplot` object can be further modified. 
- `na.omit()` returns the object with incomplete cases/rows removed. 

# Create a Scatter plot
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center', tidy=FALSE}
myTheme <- theme_bw() +
  theme(plot.title = element_text(size=8, hjust = 0.5),
        axis.title=element_text(size=8),
        legend.title = element_text(size=8),
        legend.text = element_text(size=8) )
```

- `theme_bw()`: pre-defined black and white theme. 
- Check out the complete list of available pre-defined themes [here](http://ggplot2.tidyverse.org/reference/ggtheme.html).
- `myTheme` object can be reused for different types of plot.


# Scatter plot
```{r, fig.height=2.5, fig.width=5, warning=FALSE, fig.align='center'}
g +  myTheme
```

# Scatter plot with a smoother
```{r, fig.height=2.5, fig.width=5, warning=FALSE, fig.align='center', message=FALSE}
g + geom_smooth() + myTheme
```

# Box plot using `geom_boxplot()`
```{r , fig.height=2, fig.width=3, warning=FALSE, fig.align='center', eval = FALSE, tidy=FALSE}
ggplot(na.omit(combined.long.df[,c("Age.group",
                                   "Cholesterol")]),
       aes(x = Age.group, y = Cholesterol)) +
  geom_boxplot()  +
  labs(title = "Cholesterol level versus Age",
      x = "Age Group", y = "Cholesterol level (mg/100ml)") + 
  myTheme
```
 
# Side-by-side box plot using `geom_boxplot()` and `fill`
```{r , fig.height=2, fig.width=3, warning=FALSE, fig.align='center', echo = FALSE}
ggplot(na.omit(combined.long.df[,c("Age.group", "Cholesterol")]),
       aes(x = Age.group, y = Cholesterol)) +
  geom_boxplot()  +
  labs(title = "Cholesterol level versus Age",
      x = "Age Group", y = "Cholesterol level (mg/100ml)") + 
  myTheme
```

# Side-by-side box plot using `geom_boxplot()` and `fill`

We can use the `fill` (or `colour`) argument to create side-by-side boxplots:
```{r , fig.height=2, fig.width=3, warning=FALSE, fig.align='center', eval = FALSE, tidy=FALSE}
ggplot(na.omit(combined.long.df[,c("Age.group", "Cholesterol", 
                                   "BMI.group")]),
       aes(x = Age.group, y = Cholesterol, fill = BMI.group)) +
  geom_boxplot()  +
  labs(title = "Cholesterol level versus Age and BMI",
      x = "Age Group", y = "Cholesterol level (mg/100ml)") + 
  myTheme
```
 
Note that `colour` argument changes the *points* or *lines* and `fill` argument changes the *interior area*.
 
# Side-by-side box plot using `geom_boxplot()` and `fill`
```{r , fig.height=2.5, fig.width=4, warning=FALSE, fig.align='center', echo = FALSE}
ggplot(na.omit(combined.long.df[,c("Age.group", "Cholesterol", "BMI.group")]),
       aes(x = Age.group, y = Cholesterol, fill = BMI.group)) +
  geom_boxplot()  +
  labs(title = "Cholesterol level versus Age and BMI",
      x = "Age Group", y = "Cholesterol level (mg/100ml)") + 
  myTheme
```

# Box plot with panels using `facet_wrap()`
```{r, fig.height=3, fig.width=4, warning=FALSE, fig.align='center'}
g <- ggplot(na.omit(combined.long.df[,c("Age.group", 
                                  "Cholesterol", "Race.group")]),
       aes(x = Age.group, y = Cholesterol)) +
  geom_boxplot()  +
  labs(title = "Cholesterol level versus Age",
      x = "Age Group", y = "Cholesterol level (mg/100ml)") 
```

# Box plot with panels using `facet_wrap()`
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center', message = FALSE}
g + facet_wrap(~Race.group) + myTheme + 
  theme(axis.text.x  = element_text(angle=30, vjust=1, hjust = 1))
```


# Histogram

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center', message = FALSE}
ggplot(combined.long.df, aes(x = Cholesterol)) +
  geom_histogram() + myTheme

```

# Histogram with wider binwidth

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
ggplot(combined.long.df, aes(x = Cholesterol)) +
  geom_histogram(binwidth = 10)+
  labs(x = "Cholesterol level (mg/100ml)") + myTheme
```

# Histogram with narrower binwidth

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
ggplot(combined.long.df, aes(x = Age)) +
  geom_histogram(binwidth = 1)+
  labs(x = "Cholesterol level (mg/100ml)")  + myTheme
```

# Density plot

We can substitute `geom_histogram()` from the previous code with `geom_density()` to create a density plot.

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
ggplot(combined.long.df, aes(x = Cholesterol)) +
  geom_density()+
  labs(x = "Cholesterol level (mg/100ml)")  + myTheme
```

# Density plot

Transparency is controlled with the `alpha` argument

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
ggplot(combined.long.df, aes(x = Cholesterol, fill = Age.group)) +
  geom_density(alpha = .4)+
  labs(x = "Cholesterol level (mg/100ml)")  + myTheme
```


# Bar chart
```{r, fig.height=2.5, fig.width=4, warning=FALSE, fig.align='center', tidy = FALSE}
ggplot(combined.df, aes(x = Age.group)) + geom_bar() + 
  myTheme
```

#  Age group by Smoking group  
```{r, fig.height=2, fig.width=5, warning=FALSE, fig.align='center'}
ggplot(na.omit(combined.df[,c("Age.group", "Smoke.group")]),
       aes(x = Age.group , fill = Smoke.group)) +
      geom_bar() + myTheme
```


# Bar chart on proportions

- The proportions add to 100% within each Smoke group (`group = Smoke.group`).

```{r, eval=FALSE}
g <- ggplot(na.omit(combined.df[,c("Age.group", "Smoke.group")]),
  aes(x = Age.group, y = ..prop.., group = Smoke.group,
      fill = Smoke.group))
g + geom_bar() + myTheme
```
\vspace{-30pt}
```{r, fig.height=2, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
g <- ggplot(na.omit(combined.df[,c("Age.group", "Smoke.group")]),
  aes(x = Age.group, y = ..prop.., group = Smoke.group,
      fill = Smoke.group))
g + geom_bar() + myTheme
```

# Bar chart on proportions
```{r, fig.height=2, fig.width=5, warning=FALSE, fig.align='center'}
g + geom_bar(position = "dodge") +
  labs( x = "Age group in years", y = "Percentage")+
  myTheme
```


# `ggplot2` with `dplyr` 

Create Bar chart of percentage of `Smoke.group` for each `Age.group`

- Create two-way frequency table for `Age.group` and `Smoke.group`.
- Create frequency table for `Age.group`.
- Join these two tables and compute the percentage using `mutate()`
- `geom_col()` plot the value in the final table.

# `ggplot2` with `dplyr` 

Create two-way frequency table for `Age.group` and `Smoke.group`.

```{r}
age.smoke.tab <-combined.df %>% 
  filter(!is.na(Smoke.group)) %>%
  group_by(Age.group, Smoke.group) %>% 
  count()
age.smoke.tab
```

# `ggplot2` with `dplyr` 

Create frequency table for `Age.group`.

```{r}
age.tab <- 
  combined.df %>% 
  filter(!is.na(Smoke.group)) %>%
  group_by(Age.group) %>% 
  count() %>% 
  rename(Total = n)
age.tab
```

# `ggplot2` with `dplyr` 

Join these two tables and compute the percentage using `mutate()`

```{r}
final.tab <- age.smoke.tab %>% 
  left_join(age.tab) %>% mutate(Prop = n/Total)
final.tab
```

# `ggplot2` with `dplyr` 

`geom_col()` plot the value in the final table.

```{r, fig.height=2, fig.width=5, warning=FALSE, fig.align='center'}
ggplot(final.tab, aes(x = Age.group, y = Prop, 
                      fill = Smoke.group)) +
  geom_col()+ labs(x = "Age group in years", 
                   y = "Percentage")+  myTheme
```


# Plot means in context

```{r}
combined.long.df %>% 
  group_by(Age.group) %>% 
  summarise(Mean = mean(Cholesterol, na.rm = TRUE))

```


- Means are all but meaningless unless they are presented in context.
- Always present with standard deviations (SDs) or standard error of means (SE) or confidence intervals.
- Plot means with 95% confidence intervals ($\pm$ 1.96 $\times$ SE).


# Calculating 95\% CIs
-  95\% CI $=$ Mean $\pm$ 1.96 $\times$ SE
- Standard Errors $= \frac{\text{Standard Deviation}}{\sqrt{\text{Sample Size}}}$
```{r}
combined.long.df %>% 
  group_by(Age.group) %>% 
  summarise(Mean = mean(Cholesterol, na.rm = TRUE), 
            SD = sd(Cholesterol, na.rm = TRUE))
```


# Calculating 95\% CIs
```{r}
combined.long.df %>% 
  group_by(Age.group) %>% 
  summarise(Mean = mean(Cholesterol, na.rm = TRUE), 
            SD = sd(Cholesterol, na.rm = TRUE), 
            N = sum(!is.na(Cholesterol)), 
            SE =  SD/sqrt(N), 
            CI.upr = Mean + 1.96 * SE, 
            CI.lwr = Mean - 1.96 * SE)


```

# Calculating 95\% CIs
```{r}
age.tab <- 
  combined.long.df %>% 
    group_by(Age.group) %>% 
    summarise(Mean = mean(Cholesterol, na.rm = TRUE), 
              SD = sd(Cholesterol, na.rm = TRUE), 
              N = sum(!is.na(Cholesterol)), 
              SE =  SD/sqrt(N), 
              CI.upr = Mean + 1.96 * SE, 
              CI.lwr = Mean - 1.96 * SE)
```


# Error-bars

```{r, eval=FALSE}
ggplot(age.tab, aes(x = Age.group, y = Mean)) + geom_point() +
     geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                   width = 0.1)+
     xlab("Age Group")+
     ylab("Cholesterol level (mg/100ml)") +
  myTheme
```

# Error-bars

```{r, echo=FALSE,  fig.align='center', fig.height=3, fig.width=4}
ggplot(age.tab,fig.height=3, fig.width=5, aes(x = Age.group, y = Mean)) + geom_point() +
     geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                   width = 0.1)+
     xlab("Age Group")+
     ylab("Cholesterol level (mg/100ml)") +
  myTheme
```


# Any interaction between Gender and Age group?
```{r}
GA.tab <- 
  combined.long.df %>% 
    group_by(Age.group, Sex) %>% 
    summarise(Mean = mean(Cholesterol, na.rm = TRUE), 
              SD = sd(Cholesterol, na.rm = TRUE), 
              N = sum(!is.na(Cholesterol)), 
              SE =  SD/sqrt(N), 
              CI.upr = Mean + 1.96 * SE, 
              CI.lwr = Mean - 1.96 * SE)
GA.tab
```


# Plotting mean $\pm$ 95% CI:
```{r, warning=FALSE, fig.align='center', eval=FALSE}
g <- ggplot(GA.tab, aes(x = Age.group, y = Mean,
                  color = Sex)) +
  xlab("Age Group")+
  ylab("Cholesterol level (mg/100ml)")

g + geom_point()+
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1) +
  myTheme

```

# Plotting mean $\pm$ 95% CI:
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
g <- ggplot(GA.tab, aes(x = Age.group, y = Mean,
                  color = Sex)) +
  xlab("Age Group")+
  ylab("Cholesterol level (mg/100ml)")

g + geom_point() +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1) +
  myTheme
```

# Side-by-side? `position_dodge()`
```{r, warning=FALSE, fig.align='center', eval=FALSE}
dodge <- position_dodge(width=0.2)

g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1, position = dodge) +
  myTheme
```

- `position_dodge()`, dodge overlapping objects side-to-side.  

# Side-by-side `position_dodge()`
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
dodge <- position_dodge(width=0.2)

g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1, position = dodge) +
  myTheme
```

# Flip ` coord_flip()`
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
dodge <- position_dodge(width=0.2)

g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1, position = dodge) + coord_flip() +
  myTheme
```

# Connect the points? `geom_path()`
```{r, warning=FALSE, fig.align='center', eval=FALSE}
g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1, position = dodge) +
  geom_path(aes(x = as.numeric(Age.group)),
            position = dodge) +
  myTheme
```

- ` as.numeric()`: converts any **R** object into numbers. 

```{r}
GA.tab$Age.group
```

```{r}
as.numeric(GA.tab$Age.group)
```

# Connect the points `geom_path()`
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1, position = dodge) +
  geom_path(aes(x = as.numeric(Age.group)),
            position = dodge) +
  myTheme
```

#Save a ggplot
```{r, eval=FALSE}
ggsave("myplot.pdf")
ggsave("myplot.png")

ggsave("myplot.pdf", width = 4, height = 4)
ggsave("myplot.pdf", width = 20, height = 20, units = "cm")
```
- Only save the last plot that you displayed. 
- Saved in the current working directory. 
- Detect the extention from the file name. 
- The default unit is in inches.

# Summary

```{r, echo = FALSE, results = "asis"}
data.frame(type = c("**Scatter plot**", "**Bar plot**",
                    "**Histogram**", "**Density Plot**", 
                    "**Box plot**", 
                     "**Line plot**", "**Errorbar**"),
           fun = c("`geom_point`", "`geom_bar`", 
                   "`geom_histogram`", "`geom_density`", 
                   "`geom_boxplot`",
                    "`geom_path`", "`geom_errorbar`")) %>% 
  kable(format = "markdown", 
        col.names = c("Plot type", "`geom` Function"))
```

- `ggplot2` Documentation:
[http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)

- cheatsheets:
[https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/)


