---
title: "Introduction to **R**"
subtitle: "Session 2 -- Descriptive statistics and Data subsetting"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
    fig_caption: true
    includes:
      in_header: ../../header.tex
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
options(width = 50)
patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)
```

# Reading data into **R** (Session 1)
```{r, eval=FALSE, echo = TRUE, tidy = FALSE}
patient.df <- read.csv("Data/Patient.csv", 
                       stringsAsFactors = FALSE)
str(patient.df)
```

```{r, eval=TRUE, echo = FALSE}
patient.df <- read.csv("..\\..\\Data\\Patient.csv", stringsAsFactors = FALSE)
str(patient.df)
```


# Descriptive statistics
Calculate the average of `Height`:

```{r, error=TRUE, echo = TRUE}
mean(Height)
```

You must tell **R** that `Height` is a variable (column) *within* `patient.df` data frame using `$` symbol, i.e.
```{r, error=TRUE, echo = TRUE}
mean(patient.df$Height)
```

There are missing values is the `Height` variable that **R** does not know what to do with. 

# Calculating averages with missing values

We can tell **R** to remove the missing values before calculating the average height:

```{r}
mean(patient.df$Height, na.rm = TRUE)
```


# Table of counts
```{r}
# One-way frequency table

table(patient.df$Sex)
```

# Table of proportions

```{r}
# Total count
total <- sum(table(patient.df$Sex))
total

# Proportions of total
table(patient.df$Sex)/total
```

# One-way frequency tables and `with()` 
Tired of typing `patient.df$` over and over again? Use the `with()` function.
```{r}
Sex.table <- with(patient.df, table(Sex))
Sex.table
total <- sum(Sex.table)
Sex.table/total
```


# One-way frequency tables and `with()` 

```{r}
#Convert to percentages
Sex.pct <- 100*Sex.table/total
Sex.pct
# Round to 1 decimal place
round(Sex.pct, 1)
```

#  Two-way frequency tables
```{r}
Sex.Race.tab <- with(patient.df, table(Sex, Race))
Sex.Race.tab
```

#  Two-way proportion tables
```{r}
# Calculate proportion with respect to 'margin' total
# margin = 1 (row total) or 2 (column total) 
perc.Sex.Race <- prop.table(Sex.Race.tab, margin = 2)
perc.Sex.Race
```

# Two-way proportion tables
```{r}
# Tabulate as percentages
round(100*perc.Sex.Race, 1)
```


# `which` individual does Smoke?
Let's use **R**'s powerful subsetting capabilities to select those cases for which the value of Smoke is 1 (i.e. Yes).
```{r}
index <- which(patient.df$Smoke == 1)
index
```

# How many are there?
```{r}
# Use length() to count the number of elements in "index".
length(index)
```

- `patient.df$Smoke == 1` gives a logical vector of `TRUE`/`FALSE` for every observation.
- `which()` gives an integer vector on the position of `TRUE`. 
- `length()` tells us how many elements there are in the vector.

#How many are there?
Another way is to sum up how many `TRUE`s in the logical vector `patient.df$Smoke == 1`,
```{r}
sum(patient.df$Smoke == 1)
```

#How many are there?
Another way is to sum up how many `TRUE`s in the logical vector `patient.df$Smoke == 1`
```{r}
sum(patient.df$Smoke == 1, na.rm = TRUE)
```

# Who are they?
```{r}
# Use square brackets to extract IDs corresponding  
# to the cases numbers contained in "index"
index <- which(patient.df$Smoke == 1)
patient.df$Patient.ID[index]
```

# Subsetting
Square brackets `[]` are used to extract subsets of data.
```{r}
# First element only
patient.df$Height[1]
# All but the first element
patient.df$Height[-1]
```

# Subsetting
Square brackets `[]` are used to extract subsets of data.
```{r}
# Elements 3 through 8
patient.df$Height[3:8]
# Elements 3 and 8
patient.df$Height[c(3,8)]
```

# Subsetting
Subsetting two-dimensional arrays, such as data frames, requires the use of **two** indices.
```{r}
#First row or record
patient.df[1, ] 
```

# Subsetting
For data frames, you need to use two indices.
```{r}
#Second column or variable
patient.df[, 2] 
```

# Subsetting
For data frames, you need to use two indices.
```{r}
#Some rows and columns
patient.df[2:5, 4:7] 
```

# Subsetting
For data frames, you need to use two indices.
```{r}
#Rows by number, columns by name
patient.df[1:10, c("Race", "Smoke")] 
```


# Missing values

- **R** reserves the object `NA` (`N`ot `A`vailable) for elements of a vector that are missing or unavailable.   
- Use of `is.na()` to search for missing values requires that they are recorded as `NA`.  
- `na` will not do because **R** is *case sensitive*! 

```{r} 
sum(is.na(patient.df$Smoke))
```

# Missing values
The default option of `table()` ignores `NA`s when constructing frequency tables. 

```{r}
table(patient.df$Smoke)
```

# Missing values
If you still want to see how many `NA`s in the frequency tables, you can change the `useNA` argument to `"always"` in `table()`.
```{r}
table(patient.df$Smoke, useNA = "always")
```

# Data Cleaning
The `ifelse()` function provides a quick way to convert the `Smoke` variable in `patient.df` from numbers to words, 
```{r, eval = FALSE}
  ifelse(test, yes, no)
```

- `test`: a logical test.
- `yes`, what happens if the test is True.
- `no`, what happens if the test is False.

Note, the `ifelse()` in **R** has the same functionality as `IF()` in Excel.

# Cleaning up the `Smoke` variable

Notice how we still need to remember if 1 or 2 denotes a smoker or non-smoker. 

```{r, tidy = FALSE}
patient.df$Smoke.group <- 
  with(patient.df, ifelse(Smoke == 1, "Yes", "No"))

table(patient.df$Smoke.group)
```

Notice now we have created a new variable `Smoke.group` in `patient.df` data frame.

# Cleaning up the `Race` variable

Remember that the `Race` variable has 3 levels (Caucasian, African, Other). We can nest an `ifelse()` statement inside another `ifelse()` statement.

```{r, tidy = FALSE}
patient.df$Race.group <- 
  with(patient.df,
       ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other")))
table(patient.df$Race.group)
```

# Our clean data set
```{r}
head(patient.df)
```

# Our clean data set
```{r}
str(patient.df)
```

# Our clean data set
Now we have created new variables, `Smoke.group` and `Race.group`, the next step is to remove the `Smoke` and `Race` variables in the `patient.df`.

```{r} 
names(patient.df)
```

# Removing the `Smoke` and `Race` variables
Now we have created new variables, `Smoke.group` and `Race.group`, the next step is to remove the `Smoke` and `Race` variables in the `patient.df`.

```{r} 
Smoke.index <- which(names(patient.df) == "Smoke")
Smoke.index

Race.index <- which(names(patient.df) == "Race")
Race.index

names(patient.df[,-c(Smoke.index, Race.index)])
```

# Removing the `Smoke` and `Race` variables with `%in%`

- `x %in% table` operator allows value matching, return a logical vector, indicating if a match was located for each element of `x`. 

- `!` reverse the logical vector, i.e. `TRUE` $\rightarrow$ `FALSE`.

```{r} 
names(patient.df)

names(patient.df) %in% c("Smoke", "Race")

!names(patient.df) %in% c("Smoke", "Race")
```

# Removing the `Smoke` and `Race` variables with `%in%`
```{r} 
names(patient.df[, !names(patient.df) %in% c("Smoke", "Race")])

patient.df <- 
  patient.df[, !names(patient.df) %in% c("Smoke", "Race")]

```


# Your turn

First to calculate `BMI`:
\[
\mathtt{BMI} = \frac{\mathtt{Weight(kg)}}{\mathtt{Height(m)}^2}
= \frac{\mathtt{Weight(pounds)}}{\mathtt{Height(inches)}^2} \times 703
\]

Then, create a `BMI.group` for `normal`, `overweight` and `obese`, where, 

- overweight is a BMI greater than or equal to 25; and
- obesity is a BMI greater than or equal to 30.


# Your turn
To calculate `BMI`:
\[
\mathtt{BMI} = \frac{\mathtt{Weight(kg)}}{\mathtt{Height(m)}^2}
= \frac{\mathtt{Weight(pounds)}}{\mathtt{Height(inches)}^2} \times 703
\]

```{r, tidy=FALSE}
patient.df$BMI <- (patient.df$Weight/patient.df$Height^2)*703
mean(patient.df$BMI, na.rm = TRUE)
```

Or, 
```{r, tidy=FALSE}
patient.df$BMI <- with(patient.df, (Weight/Height^2)*703)
mean(patient.df$BMI, na.rm = TRUE)
```

# Your turn

```{r, tidy=FALSE}
patient.df$BMI.group <- 
  ifelse(patient.df$BMI >= 30, "obese", 
         ifelse(patient.df$BMI >= 25, "overweight",
          "normal"))
table(patient.df$BMI.group)
table(patient.df$BMI.group, useNA = "always")

```

# Subsetting in calculations
```{r}
with(patient.df, table(Race.group, BMI.group))
```

# Subsetting in calculations
Produce the last table with Race groups of `African` and `Caucasian`, and BMI groups of `obese` and `overweight` 
```{r, tidy = FALSE}
exclude.rows <-
  with(patient.df,
     which(Race.group == "Other" | BMI.group == "normal"))
```
These are rows of Race group is `other` **or** BMI.group is `normal`

# Subsetting in calculations
Produce the last table with Race groups of `African` and `Caucasian`, and BMI groups of `obese` and `overweight` 
```{r, tidy = FALSE}
head(patient.df[-exclude.rows, c("Race.group", "BMI.group")])
```

# Subsetting in calculations
```{r, tidy=FALSE}
r.b.tab <- with(patient.df[-exclude.rows,], 
                table(Race.group, BMI.group))
r.b.tab
```

# Subsetting in calculations
Convert counts to percentages rounded to 1 decimal place.
```{r}
round(prop.table(r.b.tab)*100, 1)
```

# Different way
```{r}
r.b.tab1 <- with(patient.df, 
                 table(Race.group, BMI.group))
dim(r.b.tab1)
row.names(r.b.tab1)
colnames(r.b.tab1)
```

# Different way
```{r}
colnames(r.b.tab1)
which(row.names(r.b.tab1) == "Other")
which(colnames(r.b.tab1) == "normal")
```

# Different way
```{r}
r.b.tab1[-3, -1]

round(prop.table(r.b.tab1[-3, -1])*100, 1)
```

# Your turn 

- Average BMI of females   
- Average BMI level of males **and** Caucasian. 
      
      + Hint: *And* operator is `&`. 


# Subsetting in calculations
```{r, tidy = FALSE}
# Average BMI level of females
with(patient.df, mean(BMI[Sex == "Female"], na.rm = TRUE))

# Average BMI level of females and Caucasian
with(patient.df,
     mean(BMI[Sex == "Male" & 
              Race.group == "Caucasian"],
          na.rm = TRUE))
```

# Summary

- Making R functions
- Subsetting vectors and datasets
- `ifelse()` function











