---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 8 -- Advanced Analysis"
author: "Statistical Consulting Centre"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output: pdf_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE, eval = FALSE)

load("../../RData/Ex5.RData")
```


# 1. Logistic regression

1. Suppose the only information available about `Mercury` concentration was that the lakes were either clean (0) or contaminated (1). Create an object called `cont`, such that `cont` $= 1$ if `Mercury` $>$ 0.53 and `cont` $= 0$ if `Mercury` $\le$ 0.53.
```{r}
cont <- ifelse(joined.long.df$Mercury > 0.53, 1, 0)
```

2. Check the one-way frequency table of `cont`.
```{r}
table(cont)
```

3. Add `cont` to `joined.long.df`.
```{r}
joined.long.df$cont <- cont
```

4. (a) Perform a logistic regression to assess how `pH` influences the probability of contamination.
```{r}
myglm <- glm(cont~pH, family = binomial, data = joined.long.df)
```
    (b) What is the overall significance of `pH`?
```{r}
anova(myglm, test = "Chisq")
```
    (c) How does a 1 unit increase in `pH` change the log odds of contamination?
```{r}
summary(myglm)
```

5. (a) Perform another logistic regression to check whether the probability of contamination depends on `Calcium` level.

```{r}
myglm1 <- glm(cont~Calcium, data = joined.long.df,
              family = binomial)
```

    (b) Check the overall significance of `Calcium`.
```{r}
anova(myglm1, test = "Chisq")

summary(myglm1)
```

    (c) Calculate the estimated probability of contamination for all three `Calcium` levels using **R** as calculator.
```{r}
exp(1.0726 )/(1 + exp(1.0726 ))

exp(1.0726-1.6116 )/(1 + exp(1.0726-1.6116 ))

exp(1.0726-2.9444)/(1 + exp(1.0726-2.9444))

```
  
    (d) Check your answer using `predict()`.
```{r}
newdata = data.frame(Calcium = c("High", "Medium", "Low"))
predict(myglm1, newdata, type = "response")
```







