---
title: "Introduction to **R**"
subtitle: "Session 6 -- Advanced Graphics"
author: "Statistical Consulting Centre"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output: pdf_document
---

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = TRUE)

library(dplyr)
library(tidyr)
library(ggplot2)

load("../../RData/Ex5.RData")
```

# 1. Box plot

Using `joined.long.df`, draw a box plot showing the distribution of \texttt{Mercury} for each of the levels of \texttt{Calcium}. The boxplot should look exactly like 

```{r, fig.align='center'}
joined.long.df %>% 
  filter(!is.na(Calcium)) %>% 
  ggplot( aes(Calcium, Mercury)) + 
    geom_boxplot()+
    labs(title ="Mercury contaminations in Florida lakes",  
          x = "Calcium level", y = "Mercury concentration") +
    theme_bw()  +
    theme(plot.title = element_text(hjust = 0.5))

```

\newpage

# 2. Scatter plot
Using `joined.long.df`, draw a scatter plot which shows the relationship between `pH` and `Mercury` for each of the `Calcium` levels. The scatter plot should look exactly the same as 

```{r, fig.align='center'}
joined.long.df %>% filter(!is.na(Calcium) & !is.na(pH)) %>% 
ggplot( aes(x = pH, y = Mercury, 
        col = Calcium, shape = Calcium)) + 
  geom_point() +
  labs(title ="Mercury contaminations in Florida lakes",  
        x = "pH value", y = "Mercury concentration") +
  theme_bw()  +
  theme(plot.title = element_text(hjust = 0.5))

```
\newpage

# 3. Bar plot

Using `joined.long.df`, draw a bar plot that shows the average `Mercury` concentration for the six combinations of `Calcium` and `pHtype`. The graph should look exactly the same as

```{r, fig.align='center'}

joined.long.df %>% 
  filter(!is.na(Calcium) & !is.na(pHtype)) %>% 
  group_by(Calcium, pHtype) %>% 
  summarise(Mean = mean(Mercury)) %>% 
  ggplot(aes(x = Calcium, y = Mean, 
                           fill = pHtype)) + 
  geom_col(position = "dodge")  +
  labs(x = "Calcium level", y = "Mean Mercury concentration") +
  theme_bw()

```

\newpage

# 4. Error-bars 
1. Using `joined.long.df`, plot the mean `Mercury` concentration for each `Calcium` level (with $\pm$ 1.96 $\times$ SE bars), as shown below

```{r, fig.align='center'}
C.df <- 
  joined.long.df %>% 
  filter(!is.na(Calcium)) %>% 
  group_by(Calcium) %>%
  summarise(Mean = mean(Mercury, na.rm = TRUE),
            SD = sd(Mercury, na.rm = TRUE),
            N = sum(!is.na(Mercury)),
            SE = SD/sqrt(N),
            CI.upr = Mean + 1.96 * SE,
            CI.lwr = Mean - 1.96 * SE)

ggplot(C.df, aes(x = Calcium, y = Mean)) + 
  geom_point() +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1)  +
  labs(x = "Calcium level", y = "Mean Mercury concentration")+
  theme_bw()

```

\newpage

2. Now reproduce Figure below which shows the mean `Mercury` concentration for each `Calcium` level (with $\pm$ 1.96 $\times$ SE bars) across three `Time` periods. 

```{r, fig.align='center'}
CA.df <- 
  joined.long.df %>% 
  filter(!is.na(Calcium), !is.na(Time)) %>% 
  group_by(Calcium, Time) %>%
  summarise(Mean = mean(Mercury, na.rm = TRUE),
            SD = sd(Mercury, na.rm = TRUE),
            N = sum(!is.na(Mercury)),
            SE = SD/sqrt(N),
            CI.upr = Mean + 1.96 * SE,
            CI.lwr = Mean - 1.96 * SE)

CA.df$Time <-  factor(CA.df$Time)

dodge <- position_dodge(width=0.2)
ggplot(CA.df, aes(x = Time, y = Mean,
                  color = Calcium)) +
  xlab("Age Group")+
  ylab("Mean total Nerdy scores") +
  geom_point(position = dodge) +
  geom_errorbar(aes(ymax = CI.upr, ymin = CI.lwr),
                width = 0.1, position = dodge) +
  geom_path(aes(x = as.numeric(Time)), 
            position = dodge)  +
  labs(x = "Time", y = "Mean Mercury concentration") +
  theme_bw()

```


\newpage

# 5. Challenge: Heatmap 

Using `joined.long.df`, draw a heatmap that shows the `Mercury` concentration  for the six combinations of `Calcium` and `pHtype`. The graph should look exactly the same as
```{r, fig.align='center'}
joined.long.df %>% 
  filter(!is.na(Calcium) & !is.na(pHtype)) %>% 
ggplot(aes(x = pHtype, y = Calcium)) + 
  geom_tile(aes(fill = Mercury)) + 
  scale_fill_gradient(low = "red", high = "green") +
  theme_bw()  +
  labs(y = "Calcium level", x = "pH level")

```

Hint: `geom_tile()` for the heatmap and `scale_fill_gradient()` for the colour. 



