---
title: "Introduction to **R**"
subtitle: "Session 3 -- Basic R programming and data manipulation"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
    highlight: tango
    fig_width: 7
    fig_height: 6
    fig_caption: true
    keep_tex: true
    includes:
      in_header: ../../header.tex
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(message = FALSE)
options(width = 50)

load("../../RData/Session3.RData")

```


# Functions

A function is a relationship between a set of inputs (arguments) and a set of outputs. E.g., the function is fed some information on which it operates, the results of which are the output.

\vspace{-30pt}

```{r, echo = FALSE, message=FALSE, fig.asp=0.5}    
require(plotrix)
plot.new()
plot.window(xlim = c(-8, 8), ylim = c(-1, 1))
draw.circle(-6, 0, 2)
draw.circle(0, 0, 2)
draw.circle(6, 0, 2)
text(c(-6, 0, 6), 0, c("Input", "Function", "Output"), cex = 1.4)
arrows(c(-3.75, 2.25), 0, c(-2.25, 3.75), 0, length = 0.1)
```
\vspace{-50pt}

This is an essential building block for the **R** package.    

# Functions

We have seen many functions, e.g. `log()`, `mean()`, `table()`, `with()`, etc.

 \vspace{-30pt}
```{r, echo = FALSE, message=FALSE, fig.asp=0.5}    
plot.new()
plot.window(xlim = c(-8, 8), ylim = c(-1, 1))
draw.circle(-6, 0, 2)
draw.circle(0, 0, 2)
draw.circle(6, 0, 2)
text(c(-6, 0, 6), 0, c("1, 2, 3, 4", "Mean", "2.5"), cex = 1.4)
arrows(c(-3.75, 2.25), 0, c(-2.25, 3.75), 0, length = 0.1)
```

# Working with functions

- Functions can be user-defined, i.e., you can write your own.
- **Output is the last line of the function.** You can use `return()` to specify the output.
- Here is a function calculates the standard error of the mean (SEM):
 \vspace{-2pt}
$$
\mathrm{\hat{SEM}} = \frac{s}{\sqrt{N}}
$$
 \vspace{-7pt}

```{r}
mystder <- function(x){
    mysd <- sd(x, na.rm = TRUE) # Calc std. deviation
    n <- length(x)              # Calc sample size
    mysd/sqrt(n)                # Definition of SEM
}
mystder(patient.df$Height)
```

- A set of user-defined functions can be bundled together into an **R** package.

# **R** Packages
- Currently, the CRAN package repository features 12,619 available packages (12 June 2018). There are about 15,310 CRAN, BioConductor and Github packages in total.
- To install packages in **R**Studio, click on the **Packages** tab in the lower right, then:

    1. Click `Install`
    2. `Install from: Repository (CRAN)`
    3. Type the name of the package in `Packages`
    4. Click **Install**

- Or, you can type `install.packages("`*package name*`")`, e.g. `install.packages("plotrix")`.
- After the installation, use `library("`*package name*`")` to load it into **R**.

Note: Installation is performed only once; however, it must be loaded (i.e. use the command `install.packages("`*package name*`")` in *every* **R** session.

# Getting data into **R**

- Base **R** includes only functions which read data sets saved in simple file formats, e.g. `csv`, `txt`, tab delimited, etc.
- What if your data was saved in another format, e.g. Excel? 
- The `readxl` package for **R** contains functions that may help, [https://cran.r-project.org/web/packages/readxl/index.html][readxl]
  
```{r, eval = FALSE}
library(readxl)
excel <- read_excel("data.xlsx", sheet = 1)
```

[readxl]: https://cran.r-project.org/web/packages/readxl/index.html

# Getting data into **R**

- What if your data was saved in another format, e.g. STATA, SPSS, or SAS? 
- The `haven` package for **R** contains functions that may help, [https://cran.r-project.org/web/packages/haven/index.html][haven]
  
```{r, eval = FALSE}
library(haven)
stata <- read_dta("data.dta")
spss <- read_sav("data.sav")
sas <- read_sas("data.sas7bdat")
sasxport <- read_xpt("data.xpt")
```
However, it is always the easiest and safest to read data into **R** from a `CSV` file.

[haven]: https://cran.r-project.org/web/packages/haven/index.html


# Install **R** Packages from Bioconductor
- To install core packages, type the following in an **R** Console panel:
```{r, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite()
```
\vspace{-15pt}

- `source()` function reads **R** code from a file or, in this case, connection.

- Install specific packages, e.g., `edgeR` and `limma`, with
```{r, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite(c("edgeR", "limma"))
```

Note: Installation is performed only once; however, it must be loaded (i.e. use the command `install.packages("`*package name*`")` in *every* **R** session.

# Install **R** Packages from Github
- To install **R** Packages from Github, you need to install `devtools` **R** package first, i.e. type following in an **R** Console panel:
```{r, eval = FALSE}
install.packages("devtools")
```

- Then, install specific packages from specific author's repositories, i.e.
```{r, eval = FALSE}
devtools::install_github("author/package")
```

- For example, with my `kcha193/infoDecompuTE` package, which exists at https://github.com/kcha193/infoDecompuTE, type the following in an **R** Console panel:
```{r, eval = FALSE}
devtools::install_github("kcha193/infoDecompuTE")
```

Note: Installation is performed only once; however, it must be loaded (i.e. use the command `install.packages("`*package name*`")` in *every* **R** session.


# Loading **R** Packages
```{r, message=TRUE, warning=TRUE}
library(dplyr)
```

\vspace{-10pt}
 
- You can still use these functions from these packages by using two colons, e.g. `stats::filter()`.
- Helps to access the exact function from that specific package.
- this notation can be efficient if you only need to use a package’s function once, e.g. `devtools::install_github()`, as it avoids loading the package with `library()`.

# Different types of data objects in **R**

**R** has 6 different data types:

- character (alphanumeric; `"hello world"`),
  e.g. `Sex`
- numeric (real or decimal; `3.14159`),
  e.g. `Patient.ID`, `Age`, `Race`, `Weight`, `Height` and `Smoke`
- integer (whole numbers; `256`)
- logical (`TRUE` or `FALSE`), e.g.
```{r}
1 == 1
2 <= 0
3 != 2
```

- factor (numeric or alphanumeric, treated as categorical)
- complex (numbers with imaginary components; `3i`)

# `factor`
  \begin{block}{What is a factor?}
    \textit{A variable which takes either qualitative values, ordinal values or a discrete set of quantitative values. The values of a factor are called its \emph{levels}.}
  \end{block}
  
Examples of factors:  

- `Gender` with 2 **qualitative** levels: `Male` and `Female`.
- `Education` with 6 **ordinal** levels: `None` $<$ `Primary compl` $<$ `Incpl secondary` $<$ `Secondary compl` $<$ `Incpl university` $<$ `University degree`.
- `Income` has 9 **quantitative** levels when the mid-values of the income ranges are used: `5000`, `12500`, `17500`, `22500`, `27500`, `35000`, `45000`, `60000` and `85000`.


# `factor`

- **R** stores two *additional* pieces of information for each factor: (1) the unique set of levels and (2) an integer value, assigned by **R**, for each unique level. 
- The integer values are assigned to factor levels so that they have an order associated with them.
- By default, the unique levels are assigned the values 1, 2,..., according to ascending alphabetical order. This is not always appropriate! 

# `factor`

```{r}
class(patient.df$Sex)

patient.df$Sex <- factor(patient.df$Sex)

class(patient.df$Sex)

levels(patient.df$Sex)
```

# `factor`

```{r}
patient.df$BMI.group <- factor(patient.df$BMI.group)
levels(patient.df$BMI.group)

```


# `factor`

```{r}
patient.df$BMI.group <- factor(patient.df$BMI.group,
               levels = c("normal", "overweight", "obese"))
levels(patient.df$BMI.group)

```

# Which other variables should also be `factors`?

```{r}
str(patient.df)
```

# Which other variables should also be `factors`?

```{r}
patient.df$Race.group <- factor(patient.df$Race.group)
patient.df$Smoke.group <- factor(patient.df$Smoke.group)
```


# Converting numbers to `factors`

```{r}
test <- factor(c(0,1,2))
test
```
- then convert back to numbers? 

# And convert `factor` back to numbers

- Need to convert it to character first, using `as.character()`, then convert back to numbers, using  `as.numeric()`. 
```{r}
test 

as.numeric(test)

as.numeric(as.character(test))
```

# Your turn (Binning ages into age groups)

- Sometimes we are interested in examining responses by age group.
- Now, assign each of the `r nrow(patient.df)` patients to one of three age groups: "Under 35", "36 to 60" and "Over 61".
- Convert `Age.group` to a `factor` with levels in ascending order.


# Your turn (Binning ages into age groups)

Assign each of the `r nrow(patient.df)` patients to one of three age groups: "Under 35", "36 to 60" and "Over 61".

```{r, tidy = FALSE}
patient.df$Age.group <- 
  ifelse(patient.df$Age<=35, "Under 35", 
         ifelse(patient.df$Age <= 60, "36 to 60", "Over 61"))
table(patient.df$Age.group)

```

# Your turn (Binning ages into age groups)

Convert `Age.group` to a `factor` with levels in ascending order.

```{r}
patient.df$Age.group <- factor(patient.df$Age.group)
table(patient.df$Age.group)

```

# Your turn (Binning ages into age groups)

Convert `Age.group` to a `factor` with levels in ascending order.

```{r}
patient.df$Age.group <- factor(patient.df$Age.group, 
                           levels = c("Under 35", "36 to 60", "Over 61"))
table(patient.df$Age.group)

```

# Other way: `if`/`else` statement 

```{r, eval = FALSE}
  if(test) {
  #  yes
  } else{
  #  no
  }
```

- `test`: a logical test.
- `yes`, what happens if the test is True.
- `no`, what happens if the test is False.


# Other way: `if`/`else` statement 

```{r, eval = FALSE}
  if(test) {
  #  yes for (test)
  } else if(test1) {
  #  no for (test) but yes for (test1) 
  } else{
  #  no for both (test) and (test1)
  }
```

# Relational data

- In general, it is rare in data analysis involves only a single table of data. 
- Examples: 
    
    + Patient information and blood test measurements
    + Experimental design and measurements from the high-throughput biological instrument

# `cholesterol.df`
Serum Cholesterol level, mg/100ml, measured on:

- `Day1` 
- `Day5`
- `Day10`

# Reading data into **R**
```{r, eval=FALSE, echo = TRUE}
setwd("your working directory")
cholesterol.df <- read.csv("Data/CholesterolNA.csv")
head(cholesterol.df)
```

```{r, eval=TRUE, echo = FALSE}
cholesterol.df <- read.csv("..\\..\\data\\CholesterolNA.csv")
head(cholesterol.df)

```

# `names()`, `dim()` and `str()` functions

```{r, eval=TRUE, echo = TRUE}
names(cholesterol.df)
dim(cholesterol.df)
str(cholesterol.df)
```

# Combining data frame by columns (`cbind()`)
```{r}
combined.df <- cbind(patient.df, cholesterol.df[,-1])
```
- Thus, the function for combining the data-frame by rows is `rbind()`.
- Make sure the dimensions are correct for two combining data-frames. 
- Also, you need to make sure each row in `patient.df` matches each row in `cholesterol.df`.


# Reading data into **R** (Again)
```{r, eval=FALSE, echo = TRUE}
setwd("your working directory")
cholesterol.df <- read.csv("Data/Cholesterol.csv")
head(cholesterol.df)
```

```{r, eval=TRUE, echo = FALSE}
cholesterol.df <- read.csv("..\\..\\data\\Cholesterol.csv")
head(cholesterol.df)
```

# `names()`, `dim()` and `str()` functions

```{r, eval=TRUE, echo = TRUE}
#Names of the variables
names(cholesterol.df)
dim(cholesterol.df)
str(cholesterol.df)
```


# Combining data frame by columns (`cbind()`)
```{r, error=TRUE}
combined.df <- cbind(patient.df, cholesterol.df)
```


# Solution 

- Combining the based on the `Patient.ID` in both `patient.df` and `cholesterol.df`.
- First thing is to make sure the `Patient.ID` are unique in both data-frames. 

# Solution

- Combining the based on the `Patient.ID` in both `patient.df` and `cholesterol.df`.
- First thing is to make sure the `Patient.ID` are unique in both data-frames. 

```{r}
sum(table(patient.df$Patient.ID) > 1)
sum(table(cholesterol.df$Patient.ID) > 1)

```

# `dplyr` **R** package

- `dplyr` **R** package provides some useful functions that correspond to the most data manipulation tasks.

```{r}
library(dplyr)
```

- Mutating joins allow you to combine variables from multiple tables, there are four common types: 
    +  `left_join()`
    +  `right_join()`
    +  `full_join()`
    +  `inner_join()`


# `dplyr` **R** package
```{r, tidy=FALSE}
x <- data.frame(key = c(1,2,3), val.x = c("x1","x2","x3"))
y <- data.frame(key = c(1,2,4), val.y = c("y1","y2","y4"))
```

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.5\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.5\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
    "\\end{minipage} 
\\end{table}"
)) 
```


# `left_join()` 
```{r, eval = FALSE, message=FALSE}
left_join(x, y, by = "key")
```

- return all rows from `x`, and all columns from `x` and `y`. 
- Rows in `x` with no match in y will have `NA` values in the new columns. 
- If there are multiple matches between `x` and `y`, all combinations of the matches are returned.

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(left_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{left join}",
        t3,
    "\\end{minipage} 
\\end{table}"
)) 
```


# `right_join()` 
```{r, eval = FALSE, message=FALSE}
right_join(x, y, by = "key")
```

- return all rows from `y`, and all columns from `x` and `y`. 
- Rows in `y` with no match in x will have `NA` values in the new columns. 
- If there are multiple matches between `x` and `y`, all combinations of the matches are returned.

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(right_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{right join}",
        t3,
    "\\end{minipage} 
\\end{table}"
))  
```

# `full_join()` 
```{r, eval = FALSE, message=FALSE}
full_join(x, y, by = "key")
```

- return all rows and all columns from both `x` and `y`. 
- Where there are not matching values, returns `NA` for the one missing.

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(full_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{full join}",
        t3,
    "\\end{minipage} 
\\end{table}"
))  
```

# `inner_join()` 
```{r, eval = FALSE, message=FALSE}
inner_join(x, y, by = "key")
```

- return all rows from `x` where there are matching values in `y`, and all columns from `x` and `y`.
- If there are multiple matches between `x` and `y`, all combination of the matches are returned.

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(inner_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{inner join}",
        t3,
    "\\end{minipage} 
\\end{table}"
)) 
```

# `dplyr` **R** package

These four types of mutating join are differ in their behavior when a match is not found. 

-  `left_join()`
-  `right_join()`
-  `full_join()`
-  `inner_join()`

Which one should we use for `patient.df` and `cholesterol.df` data-frames?

# Combining two data frames

```{r, message=TRUE}
combined.df <- left_join(patient.df, cholesterol.df)
```

 
```{r}
combined.df <- left_join(patient.df, cholesterol.df, 
                         by =  "Patient.ID")
```

- Variable names:
```{r}
names(combined.df)
```

# Combining two data frames

```{r}
str(combined.df)
```

# Renaming variables
You can change the variable names.
```{r, tidy=FALSE}
names(combined.df)[c(1,11,12,13)] <- 
  c("ID", "Baseline", "PreTrt", "PostTrt")

names(combined.df)
```

# Summary

- functions
- **R** packages
- `factor`
- Combining two data frames











