---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 7 -- Simple Analysis"
author: "Statistical Consulting Centre"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output: pdf_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE, eval = FALSE)

load("../../RData/Ex5.RData")

```


# 1. T-test

1. Do a two-sample $t$-test (two-sided) to see whether `Mercury` for acidic lakes is different from other `pHtype` levels, assuming 

    (a) population variances are equal, and

```{r}
joined.long.df$pHtypeAcidic <- with(joined.long.df, ifelse(pH < 7, "acidity", "Others"))

with(joined.long.df, t.test(Mercury~pHtypeAcidic, var.equal = TRUE))
```

    (b) population variances are not equal

```{r}

with(joined.long.df, t.test(Mercury~pHtypeAcidic, var.equal = FALSE))
```

Hint: Create a new variable `pHtypeAcidic` for just acidic group.

2. How do the results from using the following code compare with the results you got in question 1?

````
acid.mer <- with(joined.long.df, mercury[pHtype == "acidic"])
not.acid.mer <- with(joined.long.df, mercury[pHtype != "acidic"])

t.test(acid.mer, not.acid.mer)

````

```{r}

acid.mer <- with(joined.long.df, Mercury[pHtype == "acidic"])
not.acid.mer <- with(joined.long.df, Mercury[pHtype != "acidic"])

t.test(acid.mer, not.acid.mer)

```


# 2. One-way ANOVA

1. Perform an ANOVA to test the null hypothesis that the mean `Mercury` concentrations are all equal for all three `Calcium` levels.

```{r}

myaov <- with(joined.long.df, aov(Mercury ~ Calcium))

```

2. Use `summary()` to check the overall signifcance of `Calcium`.

```{r}

summary(myaov)

```


3. Use `model.tables()` to calculate the mean `Mercury` concentrations for each `Calcium` level.


```{r}
model.tables(myaov, type = "means")
```


4. Carry out pairwise comparisons of mean `Mercury` concentrations between pairs of `Calcium` levels, adjusting the $p$-values using Tukey's Honestly Signifcant Difference test.

```{r}
TukeyHSD(myaov)

```


# 3. Two-way ANOVA

1. Perform a Two-way ANOVA to test the null hypothesis that the mean `Mercury` concentrations are all equal for all three `Calcium` levels across all three days.

```{r}

myaov <- with(joined.long.df, aov(Mercury ~ Calcium*Time))

```

2. Use `summary()` to check the overall signifcance.

```{r}

summary(myaov)

```


3. Use `model.tables()` to calculate the mean `Mercury` concentrations for each `Calcium` level.


```{r}
model.tables(myaov, type = "means")
```


4. Carry out pairwise comparisons of mean `Mercury` concentrations, adjusting the $p$-values using Tukey's Honestly Signifcant Difference test.

```{r}
TukeyHSD(myaov)

```


# 4. Test of independence

1. Perform a Pearson's Chi-squared test to check if `Calcium` depend on `pHtype` level? 

```{r}
ca.ph.tab <- with(joined.df, table(Calcium, pHtype))
ca.ph.tab

chisq.test(ca.ph.tab)

```

2. Why is there a warning from 4.1? What is an altertive test which we can perform?


```{r} 
fisher.test(ca.ph.tab)

```


# 5. Linear regression

1. Produce a scatter plot between `pH` (x-axis) and `Mercury` (y-axis).

```{r}
with(joined.long.df, plot(pH, Mercury))

```


2. Describe the relationship between `pH` and `Mercury` shown in the scatter plot.


3. Fit a linear model between `pH` and `Mercury`, in which `Mercury` is the dependent (response) variable and `pH` is the independent (explanatory) variable. Name the linear model as `mylm`.


```{r}
mylm <- with(joined.long.df, lm(Mercury ~ pH))
```

4. Use `summary()` to obtain the estimates and test the statistical signifcance of the intercept and slope.

```{r}
summary(mylm)
```

5. Write down the equation of the fitted line.


6. Use the following code to check the residuals, for homogeneity of variance.

````
plot(predict(mylm), residuals(mylm))
abline(h = 0, lwd = 2, col = 2)
````

```{r}
plot(predict(mylm), residuals(mylm))
abline(h = 0, lwd = 2, col = 2)
```








