---
title: "Introduction to **R**"
subtitle: "Session 7 -- Simple analysis"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output:
  beamer_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: TRUE
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
    highlight: tango
    includes:
      in_header: ../../header1.tex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, small.mar=TRUE, message = FALSE)

options(width = 100)
library(dplyr)
library(tidyr)
library(ggplot2)

load("../../Rdata/Session5.Rdata")
```


# Regression functions

Two of the most commonly used **R** commands for modeling:

- `lm()`: fits **L**inear **M**odels
- `glm()`: fits **G**eneralised **L**inear **M**odels.

Note SAS users: `PROC GLM` is **not** the same as **R**'s `glm()`.

There's a lot in these two commands; entire stage 3 statistical courses on linear and generalised linear models.

# Student's T-test in **R**

```
t.test(y ~ x, data = dataset)
```

`y`: the continuous response variable.

`x`: grouping variable with 2 levels.

`data`: name of the data frame containing the variables.


- Suppose we want to test whether males and females (`x = Sex`) have different Cholesterol levels.
- After visualising the data, we can perform the t-test in **R**:
- Categorical variables should be converted to type `factor` before analysis, i.e.
```
combined.long.df$Sex <- factor(combined.long.df$Sex)
```

# Student's t-test in **R**

```{r, echo = FALSE,  warning = FALSE, fig.align='center', out.width = "75%"}

ggplot(combined.long.df, aes(Sex, Cholesterol)) +
  geom_boxplot() +
  theme_bw()

```

# Student's t-test in **R**

\vspace{-5pt}

Note that we could log-transform the cholesterol variable and make inferences on the median.

\vspace{-10pt}

```{r, echo = FALSE,  warning = FALSE, fig.align='center', out.width = "70%"}
ggplot(combined.long.df, aes(Sex, log(Cholesterol))) +
  geom_boxplot() +
  labs(y = "Log cholesterol") +
  theme_bw()
```

# Student's t-test in **R**

```{r}
t.test(Cholesterol ~ Sex, data = combined.long.df)
```

- $p$-value = `r round(t.test(Cholesterol ~ Sex, data = combined.long.df)$p.value, 4)`.
- We have a strong evidence that the average cholesterol level for male is between 2.95 and 17.5 mg/100ml higher than for female.


# Analysis of Variance (ANOVA) in **R** 

- Generalises the t-test to more than 2 groups.
- Null hypothesis: all group means are equal.
```
aov(y ~ x, data = dataset)
```

# Analysis of Variance (ANOVA) in **R** 

```{r, echo = FALSE,  warning = FALSE, fig.align='center', out.width = "75%"}

ggplot(combined.long.df, aes(Age.group, Cholesterol)) +
  geom_boxplot() +
  theme_bw()

```

# Analysis of Variance (ANOVA) in **R** 
**Example.** Mean Cholesterol level is the same for all three Age groups

```{r}
tryaov = aov(Cholesterol ~ Age.group, data = combined.long.df)
```

- `aov()`: **A**nalysis **o**f **V**ariance.
- Response variable (i.e. `Cholesterol`) is separated by `~` from explanatory variable(s) (i.e. `age.group`).
- All explanatory variables should be categorical (otherwise it's not ANOVA).

# Analysis of Variance (ANOVA) in **R** 

```{r}
summary(tryaov)
```

We have extremely strong evidence that the average cholesterol level in at least one age group is different to at least one other age group.

Which one(s) is(are) different????

# Group means

We can compute a summary table of the results easily with the `model.tables` function:

```{r}
model.tables(tryaov, "means")
```

```{r echo = FALSE}
m <- round(model.tables(tryaov, "means")$tables$'Grand mean', 1)
gm <- round(model.tables(tryaov, "means")$tables$Age.group, 1)
```

The mean Cholesterol level...

- over all participants is `r m`.


# Group means
```{r}
model.tables(tryaov, "means")
```

```{r echo = FALSE}
m <- round(model.tables(tryaov, "means")$tables$'Grand mean', 1)
gm <- round(model.tables(tryaov, "means")$tables$Age.group, 1)
```

The mean Cholesterol level...

- for "Under 35" group is lower than both that of the "36 to 60" and the "Over 61" groups.
- for "36 to 60" group is lower than the "Over 61" group.


# Group means
```{r}
model.tables(tryaov, "means")
```

Are any pairs of these means statistically different from one another?

# Post-hoc multiple comparisons
\vspace{-4mm}
```{r}
TukeyHSD(tryaov)
```
\vspace{-6mm}

- `diff`: estimated difference between two group means.
- `lwr, upr`: lower and upper limit of the 95\% confidence interval of the estimated difference.
- `p adj`: `p`-values adjusted for multiple comparisons.


# Post-hoc multiple comparisons
```{r}
comp <- TukeyHSD(tryaov)
comp$Age.group
```

- Mean Cholesterol level for "36 to 60" is `r abs(round(comp$Age.group[1, 1], 1))` mg/100ml **higher** than "Under 35" (`p adj` $<$ 0.0001).
- Mean Cholesterol level for "Over 61" is `r abs(round(comp$Age.group[2, 1], 1))` mg/100ml **higher** than "Under 35" (`p adj` $<$ 0.0001).
- Mean Cholesterol level for "Over 61" is `r abs(round(comp$Age.group[3, 1], 1))` mg/100ml **higher** than "36 to 60" (`p adj` $<$ 0.0001).

# From Session 6: Mean Cholesterol level vs Age group

```{r, echo = FALSE, fig.align='center', out.width = "75%"} 
pp <- data.frame(comp$Age.group)
pp$Age.group <- factor(rownames(comp$Age.group))

library(ggplot2)
ggplot(pp, aes(x = Age.group, y = diff)) + geom_point() + 
  geom_errorbar(aes(ymax = upr, ymin = lwr), width = 0.25) + 
    theme_bw() +
  labs(x = "Age group", y = "Difference",
       title = "Tukey HSD intervals for each age group comparision") +
  theme(text = element_text(size = 14))
```



# Two-way ANOVA in **R**

- The last ANOVA model was fitted using on categorical variable (`Age.group`), hence a *one*-way ANOVA.
- If we fit a linear model using two categorical, explanatory variables, we have a *two*-way ANOVA.
- Recall: All categorical variables should be converted into `factor`s.


# Two-way ANOVA in **R**

```{r, echo = FALSE,  warning = FALSE, fig.align='center', out.width = "75%"}

ggplot(combined.long.df, aes(Age.group, Cholesterol)) +
  geom_boxplot() +
  facet_wrap(~Sex) +
  theme_bw()

```

# Two-way ANOVA in **R**

```{r, tidy = FALSE}
try2way <-  aov(Cholesterol ~ Sex * Age.group, combined.long.df)
```
  
- `Sex * Age.group` is equivalent to `Sex + Age.group + Sex:Age.group`.


# Two-way ANOVA in **R**
```{r}
summary(try2way)
```

There is significant two-way interaction between \texttt{Sex} and \texttt{Age.group} ($p$-value = 0.0178), i.e., the magnitude of the difference in mean Cholesterol levels between males and females is constant across all age groups, and vice versa.

# Estimated means
```{r}
model.tables(try2way, "means")
```

# Post-hoc pairwise comparisons
```{r}
TukeyHSD(try2way)$`Sex:Age.group`
```

# Three-way ANOVA with `Time` in **R**
```{r}
try3way <-  aov(Cholesterol~Sex*Age.group*Time, data = combined.long.df)
               

summary(try3way)
```


# Test of independence in **R**

```{r, tidy=FALSE}
smoke.age.tab <- with(combined.df, table(Smoke.group,
                                         Age.group))
smoke.age.tab
```

Do smoking habit depend on age group? 

Statistically speaking, is `Smoke.group` and `Age.group` independent of one another?


# Pearson's Chi-squared test in **R**

We can use the `chisq.test` function in **R** to perform a Pearson's Chi-square test for independence:

```{r}
chisq.test(smoke.age.tab)
```

- There is evidence ($p$-value $=$ 0.0011) that `Smoke.group` and `Age.group` are not independent of one another.

- Smoking habit does depend on the age group to which patient belong.


# Assumptions

- Pearson's Chi-squared tests have certain assumptions.
- These assumptions are primarily to do with sample size.
- **R** will give you a warning if these assumptions are not met:

```{r echo = FALSE, error=TRUE}
mytest <- 1:4
chisq.test(mytest)
```

- These assumptions are more likely to be wrong if the sample size is small.
- If this happens, the alternative is to use Fisher's exact test.


# Fisher's exact test in **R**
Assume `smoke.age.tab` does not meet the underlying assumptions of Pearson's Chi-squared test.

```{r}
fisher.test(smoke.age.tab)
```

# Linear regression in **R**

`lm(y ~ x, data = dataset)` is used for linear regression.

`y`, the response variable.

`x`, the explanatory variable.   

`data`: name of the data frame containing the variables.

- There can be more than one explanatory variable, called **multiple** linear regression. 

# Simple linear regression in **R**

When there is only one explanatory variable here, we refer to this a **simple ** linear regression.

```{r, tidy = FALSE, fig.align='center', eval = FALSE}
with(combined.long.df, plot(Age, Cholesterol))
```
\vspace{-35pt}
```{r, fig.align='center', echo = FALSE, out.width = "70%"}
with(combined.long.df, plot(Age, Cholesterol))
```


# Fit the regression model 

Let’s carry out the linear regression of Age on Cholesterol level, i.e.
```{r}
trylm <- lm(Cholesterol ~ Age, data = combined.long.df)
```

We now need to check our assumption that the residuals are normally distributed.

# Check normality of the residuals

- We can extract the residuals of the model with `resid()`.
- We can plot a Quantile-Quantile (QQ) plot with ``qqnorm()`

```{r, eval=FALSE, fig.align='center'}
qqnorm(resid(trylm))
qqline(resid(trylm), col = "red")
```
\vspace{-30pt}
```{r, fig.align='center', echo = FALSE, out.width = "60%"}
qqnorm(resid(trylm))
qqline(resid(trylm), col = "red")
```

# Check Residuals plot

```{r, fig.align='center', eval=FALSE}
plot(predict(trylm), residuals(trylm))
abline(h = 0, col = 2, lwd = 2)
```
\vspace{-30pt}
```{r, fig.align='center', echo=FALSE, out.width = "75%"}
plot(predict(trylm), residuals(trylm))
abline(h = 0, col = 2, lwd = 2)
```

# Check normality of the residuals

We can refit the model using log cholesterol as our response variable instead:

```{r, eval=FALSE}
trylmLog <- lm(log(Cholesterol) ~ Age, data = combined.long.df)
qqnorm(resid(trylmLog))
qqline(resid(trylmLog), col = "red")
```
\vspace{-30pt}
```{r,  fig.align='center', echo = FALSE, out.width = "65%"}
trylmLog <- lm(log(Cholesterol) ~ Age, data = combined.long.df)
qqnorm(resid(trylmLog))
qqline(resid(trylmLog), col = "red")
```
# Check Residuals plot

```{r, fig.align='center', eval=FALSE}
plot(predict(trylmLog), residuals(trylmLog))
abline(h = 0, col = 2, lwd = 2)
```
\vspace{-30pt}
```{r, fig.align='center', echo=FALSE, out.width = "75%"}
plot(predict(trylmLog), residuals(trylmLog))
abline(h = 0, col = 2, lwd = 2)
```

# Simple linear regression in **R**

```{r, tidy = FALSE, eval=FALSE}
summary(trylm)
```

\vspace{-20pt}

```{r, echo = FALSE}
summary(trylm)
```

# Simple linear regression in **R**

```{r, tidy = FALSE, eval=FALSE}
summary(trylmLog)
```

\vspace{-20pt}

```{r, echo = FALSE}
summary(trylmLog)
```

# Final regression model

```{r, echo = FALSE}
round(summary(trylmLog)$coef, 4)
```


- The estimated intercept is 5.0765 There is a very strong evidence that this is not zero ($p$-value < 0.0001). 
- The estimated slope of `Age` is 0.0042. There is a very strong evidence that this is not zero ($p$-value < 0.0001).
- The fitted line is:

$$ log(Cholesterol) = 5.0765 + 0.0042 \times Age $$
$$ Cholesterol = e^{5.0765 + 0.0042 \times Age} $$

- For every one year increase in age, the Cholesterol level increase by (exp(0.0042)) = `r exp(0.0042)` **times**.  


# Conclusion

- The linear relationship between age and Cholesterol level is statistically significant.
- Average cholesterol level is positive related to age.

# Add the fitted line

```{r, warning=FALSE, fig.align='center', eval=FALSE}
with(combined.long.df, plot(Age, log(Cholesterol)))
abline(trylmLog, col = 2, lwd = 2)
```
\vspace{-30pt}
```{r, warning=FALSE, fig.align='center', echo=FALSE, out.width = "75%"}
with(combined.long.df, plot(Age, log(Cholesterol)))
abline(trylmLog, col = 2, lwd = 2)
```

# Quadratic term

```{r, tidy=FALSE}
tryquad <- lm(log(Cholesterol) ~ Age + I(Age^2), data = combined.long.df)
round(summary(tryquad)$coef, 4)
```

`I(Age^2)` tells R to treat `^` as arithmetical operator, rather than formula operator.

Our fitted curve is:

$$ log(Cholesterol) = 4.7114 + 0.0206 \times Age -0.0002 \times Age^2 $$

# Summary


```{r, echo = FALSE}
data.frame(model = c("**Student's t-test**", "**One-way ANOVA**",
                    "**Two-way ANOVA**", "**Pearson's Chi-square test**",
                    "**Fisher's exact test**", "**Linear regression**"),
           fun = c("`t.test()`", "`aov()`", "`aov()`", "`chisq.test()`",
                   "`fisher.test()`", "`lm()`")) %>% 
  knitr::kable(format = "markdown", 
        col.names = c("Model", "Function"))
```



