---
title: "Introduction to **R**"
subtitle: "Session 4"
author: "Statistical Consulting Centre"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output: pdf_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
library(dplyr)
library(tidyr)

load("../../RData/Ex4.RData")


```


# 1. `for` loop

1. Check the one-way frequency tables of pHtype and Calcium.

```{r}
for (i in c("pHtype", "Calcium")){
  print(i)
  print(table(joined.df[,i]))
}
```


2. Calculate the means and standard deviations of pH, Chlorophyll and mercury measuremnts at Day 1, 2, and 3. 

```{r}
for (i in c("pH", "Chlorophyll", "Day1", "Day2", "Day3")){
  print(i)
  print(mean(joined.df[,i], na.rm = TRUE))
  print(sd(joined.df[,i], na.rm = TRUE))
}
```


# 2. `apply()` and `tapply()` functions

1. Check the one-way frequency tables of `pHtype` and `Calcium`.

```{r}
apply(joined.df[, c("pHtype", "Calcium")], 2, table)

```

2. Calculate the means and standard deviations of `pH`, `Chlorophyll` and mercury measurements at Day 1, 2, and 3 using `apply()`.

```{r}
apply(joined.df[, c("pH", "Chlorophyll", "Day1", "Day2", "Day3")], 2, mean, na.rm=TRUE)

apply(joined.df[, c("pH", "Chlorophyll", "Day1", "Day2", "Day3")], 2, sd, na.rm=TRUE)

```

3. Now calculate the means and standard deviations of pH, Chlorophyll and mercury to only 2 decimal places using `apply()`.

```{r}
round(apply(joined.df[, c("pH", "Chlorophyll", "Day1", "Day2", "Day3")], 2, mean, na.rm = TRUE), 2)

round(apply(joined.df[, c("pH", "Chlorophyll", "Day1", "Day2", "Day3")], 2, sd, na.rm = TRUE), 2)

```

4. How many non-missing values are there in each of the variables, in `joined.df`?

```{r}

apply(joined.df[, -c(1,2)], 2, function(x) sum(!is.na(x)))

```


5. You were introduced to the following function, `mytab()`, in the Session 4 lecture slides.


`````
mytab <- function(someinput){
  n <- length(someinput)
  n.missing <- na.check(someinput)
  n.complete <- n - n.missing
  mymean <- round(mean(someinput, na.rm = TRUE), 2)
  mysd <- round(sd(someinput, na.rm = TRUE), 2)
  mystder <- round(mysd/sqrt(n.complete), 2)
  Lower.CI <- round(mymean - 1.96*mystder, 2)
  Upper.CI <- round(mymean + 1.96*mystder, 2)
  
  c(Complete.obs = n.complete, Missing.obs = n.missing,
  Mean = mymean, Std.Error = mystder,
  Lower.CI = Lower.CI, Upper.CI = Upper.CI)
}
`````


It depends on the `na.check()` function, defined earlier, to calculate the number of missing values, i.e., `mytab()` depends on the availability of `na.check()` in order for it to work. Modify `mytab()` so it no longer depends on `na.check()` to calculate the number of missing values. Let's call the modified function `mytab1()`.


```{r}
mytab1 <- function(someinput){
  n <- length(someinput)
  n.missing <- length(which(is.na(someinput)))
  n.complete <- n - n.missing
  mymean <- round(mean(someinput, na.rm = TRUE), 2)
  mysd <- round(sd(someinput, na.rm = TRUE), 2)
  mystder <- round(mysd/sqrt(n.complete), 2)
  Lower.CI <- round(mymean - 1.96*mystder, 2)
  Upper.CI <- round(mymean + 1.96*mystder, 2)
  c(Complete.obs = n.complete, Missing.obs = n.missing,
  Mean = mymean, Std.Error = mystder,
  Lower.CI = Lower.CI, Upper.CI = Upper.CI)
}


```

6. Use `mytab1()` to produce a summary table for all continuous variables in `joined.df`.

```{r}
apply(joined.df[, c("pH", "Chlorophyll", "Day1", "Day2", "Day3")], 2, mytab1)

```


7. Use `mytab1()` to produce a summary table of `pH` for all `Calcium` levels in `joined.df`.

```{r}
with(joined.df, tapply(pH, Calcium, mytab1))

```


# 3. `tidyr` **R** package

Create a new data set `joined.long.df` using `gather()` function in `tidyr` **R** package to combine the measurements at Day 1, 2, and 3 for the `Mercury` variable.  


```{r} 
library(tidyr)
joined.long.df <- gather(joined.df, "Day1", "Day2", "Day3", key = Time, value= Mercury)

```



# 4. Challenge 

1. Use `%>%` piping operator with functions in `dplyr` and `tidyr` **R** packages to recreate `joined.df` and then `joined.long.df`.


```{r, eval = FALSE, tidy = FALSE}
lake.df <- read.csv("location of your folder/Lake.csv", 
                      stringsAsFactors = FALSE)

mercury.df <- read.csv("location of your folder/Mercury.csv", 
                      stringsAsFactors = FALSE)

joined.df <- lake.df %>% 
  mutate(pHtype = factor(ifelse(pH > 7, "alkalinity", 
                                ifelse(pH < 7, "acidity", "natural")), 
                         levels = c( "acidity", "natural", "alkalinity")), 
         Calcium = factor(lake.df$Calcium, 
                          levels = c("Low", "Medium", "High"))) %>% 
  left_join(mercury.df, by = "ID")


joined.long.df <- joined.df %>% gather("Day1", "Day2", "Day3", 
                                       key = Time, value= Mercury)

```


```{r, echo = FALSE}

lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)
mercury.df <- read.csv("..\\..\\data\\Mercury.csv", 
                      stringsAsFactors = FALSE)

joined.df <- lake.df %>% 
  mutate(pHtype = factor(ifelse(pH > 7, "alkalinity", ifelse(pH < 7, "acidity", "natural")), 
                                levels = c( "acidity", "natural", "alkalinity")), 
         Calcium = factor(lake.df$Calcium, levels = c("Low", "Medium", "High"))) %>% 
  left_join(mercury.df, by = "ID")


joined.long.df <- joined.df %>% gather("Day1", "Day2", "Day3", key = Time, value= Mercury)

```

2. Use `group_by()` and `summarise()` functions to calculate the mean of mercury to only 2 decimal places for all six combinations of `pHtype`, `Calcium` and `Time`.

```{r}

joined.long.df %>% group_by(pHtype, Calcium, Time) %>% 
  summarise(Mean = round(mean(Mercury, na.rm = TRUE), 2))

```
