---
title: "Introduction to **R**"
subtitle: "Session 2"
author: "Statistical Consulting Centre"
date:  |
  | `r readLines('../../metadata.txt')[2]`
output: pdf_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE, eval = FALSE)

lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)

```

# 1. Descriptive statistics

1. Calculate the mean and standard deviation of both `pH` and `Chlorophyll`.

```{r}
mean(lake.df$pH, na.rm = TRUE)
mean(lake.df$Chlorophyll, na.rm = TRUE)

sd(lake.df$pH, na.rm = TRUE)
sd(lake.df$Chlorophyll, na.rm = TRUE)

```

2. Check out what `summary()` does by running `summary(lake.df$pH)`.

```{r}
summary(lake.df$pH)

```

3. Check the frequency of `Calcium` concentration.

```{r}
table(lake.df$Calcium)

```

4. Turn the frequency table from above into a table of proportions, keep only 2 decimal places.

```{r}
round(prop.table(table(lake.df$Calcium)) * 100, 2)

```


# 2. Subsetting

1. Print the following to the console:
    
    - The `pH` of the first lake.
    
```{r}
lake.df$pH[1]
```

    - The `pH` of the last lake.
    
```{r}
lake.df$pH[53]
#or:
lake.df$pH[nrow(lake.df)]
```

    - The `pH` values of the first and last lakes.
    
```{r}
lake.df$pH[c(1, 53)]
```

    - All measurements made on the third lake.
    
```{r}
lake.df[3, ]
```

    - All `pH` values.
    
```{r}
lake.df[, "pH"]
```

2. Calculate:

    - The average `pH` of lakes with low `Calcium` concentrations.
    
```{r}
with(lake.df, mean(pH[Calcium == "Low"], na.rm = TRUE))
```

    - The average `pH` of lakes with low `Calcium` concentrations and `Chlorophyll` concentrations lower than 10.
    
```{r}
with(lake.df, mean(pH[Calcium == "Low" & Chlorophyll < 10], na.rm = TRUE))
```

# 3. `ifelse()` function

1. Create a new variable called `pHtype` which has the value `acidic` if `pH` $< 7$, the value `basic` if `pH` $> 7$ and the value `natural` if `pH` $== 7$.

```{r}
lake.df$pHtype <- with(lake.df, ifelse(pH > 7, "basic", ifelse(pH < 7, "acidic", "natural")))

```

2. Produce a two-way frequency table of `Calcium` versus `pHtype`.

```{r}
with(lake.df, table(Calcium, pHtype))

```